An Evaluation of Severe Weather Data from 1950 - 2011
==============================================================================
## Synopsis: 
### This report evaluates weather data acquired over 60 years (1950-2011) by the National Oceanic and Atmospheric Administration (NOAA). The analysis is narrowly focused on determining (1) the weather events that are most harmful to human health and (2) which weather events have the greatest ecomonic consequences across the United States. The analysis shows that Tornados are the largest source of human casualties (including injuries and fatalities) from weather related activity, accounting for 64% of casualties over the sampling period. Furthermore, the analysis indicates that flooding is the reason for most of the weather-related economic loss in half of all the 50 states.

## Data Processing: 
###The NOAA database was downloaded from web via the John Hopkins DataScience-Coursera site and stored on the computer with which the analysis was done. The .csv file (repdata-data-StormData.csv) consisting of 902297 rows and 37 columns, was then loaded into the R environment as follows:

```{r,echo=TRUE}
setwd("C://Users//solasoy//Dropbox//R//Coursera//Reproducible Research")
x <- read.csv(file='repdata-data-StormData.csv',
              colClasses="character")
```
###The 985 weather events that were recorded over time were found to contain repititions and other redundant information. Accordingly the data required some filtering followed by the categorization of weather activity into distinct groups to facilitate further analysis. THe NOAA has identified 48 groups of weather event categories (pg 6 of the cheat sheet provided by the NOAA). To the extent that it was feasible, all the listed wheather events were grouped into these categories and the rest removed, yielding a final weather weather event count of 656 or 890365 rows of data.


### The data pre-processing step was carried out as follows:

```{r, Data cleaning,echo=TRUE}
x$INJURIES <- as.numeric(x$INJURIES)
x$FATALITIES <- as.numeric(x$FATALITIES)

u <- unique(x$EVTYPE)

# Consolidate event types 48 categories defined by NOAA
# First 24 ---
a1 <- union(u[grepl("Tide",u,ignore.case=TRUE)],u[grepl("Astronomical",u,ignore.case=TRUE)])
a2 <- u[grepl("Avalanche",u,ignore.case=TRUE)]
a3 <- u[grepl("Blizzard",u,ignore.case=TRUE)]
a4 <- u[grepl("Coastal Flood",u,ignore.case=TRUE)]
a5 <- union(u[grepl("Cold Chill",u,ignore.case=TRUE)],u[grepl("Wind Chill",u,ignore.case=TRUE)])
a5 <- a5[!grepl("Extreme",a5,ignore.case=TRUE)]
a6 <- u[grepl("Debris Flow",u,ignore.case=TRUE)]
a7 <- u[grepl("Dense Fog",u,ignore.case=TRUE)]
a8 <- u[grepl("Dense Smoke",u,ignore.case=TRUE)]
a9 <- u[grepl("Drought",u,ignore.case=TRUE)]
a10 <- u[grepl("Dust Devil",u,ignore.case=TRUE)]
a11 <- u[grepl("Dust Storm",u,ignore.case=TRUE)]
a12 <- u[grepl("Excessive Heat",u,ignore.case=TRUE)]
a13 <- union(u[grepl("Extreme Cold Chill",u,ignore.case=TRUE)],u[grepl("Extreme Wind Chill",u,ignore.case=TRUE)])
a14 <- u[grepl("Flash Flood",u,ignore.case=TRUE)]
a15 <- u[grepl("Flood",u,ignore.case=TRUE)]
a15 <-  a15[!grepl("Lakeshore",a15,ignore.case=TRUE)]
a16 <- union(u[grepl("Frost",u,ignore.case=TRUE)],u[grepl("Freeze",u,ignore.case=TRUE)])
a17 <- u[grepl("Funnel Cloud",u,ignore.case=TRUE)]
a18 <- u[grepl("Freezing Fog",u,ignore.case=TRUE)]
a19 <- u[grepl("Hail",u,ignore.case=TRUE)]
a19 <-  a19[!grepl("Marine",a19,ignore.case=TRUE)]
a20 <- u[grepl("Heat",u,ignore.case=TRUE)]
a21 <- u[grepl("Heavy Rain",u,ignore.case=TRUE)]
a22 <- u[grepl("Heavy Snow",u,ignore.case=TRUE)]
a23 <- u[grepl("High Surf",u,ignore.case=TRUE)]
a24 <- u[grepl("High Wind",u,ignore.case=TRUE)]

# Last 24......
b1 <- union(u[grepl("Hurricane",u,ignore.case=TRUE)],u[grepl("Typhoon",u,ignore.case=TRUE)])
b2 <- u[grepl("Ice Storm",u,ignore.case=TRUE)]
b3 <- u[grepl("Lake effect snow",u,ignore.case=TRUE)]
b4 <- u[grepl("Lakeshore Flood",u,ignore.case=TRUE)]
b5 <- union(u[grepl("Lightning",u,ignore.case=TRUE)],u[grepl("Wind Chill",u,ignore.case=TRUE)])
b6  <- u[grepl("Marine Hail",u,ignore.case=TRUE)]
b7 <- u[grepl("Marine High Wind",u,ignore.case=TRUE)]
b8 <- u[grepl("Marine Strong Wind",u,ignore.case=TRUE)]
b9 <- u[grepl("Marine Thunderstorm Wind",u,ignore.case=TRUE)]
b10 <- u[grepl("Rip Current",u,ignore.case=TRUE)]
b11 <- u[grepl("Sieche",u,ignore.case=TRUE)]
b12 <- u[grepl("Sleet",u,ignore.case=TRUE)]
b13 <- u[grepl("Storm Surge/Tide",u,ignore.case=TRUE)]
b14 <- u[grepl("Strong Wind",u,ignore.case=TRUE)]
b15 <- union(u[grepl("Thunderstorm Wind",u,ignore.case=TRUE)],u[grepl("TSTM Wind",u,ignore.case=TRUE)])
b16 <- u[grepl("Tornado",u,ignore.case=TRUE)]
b17 <- u[grepl("Tropical Depression",u,ignore.case=TRUE)]
b18 <- u[grepl("Tropical Storm",u,ignore.case=TRUE)]
b19 <- u[grepl("Tsunami",u,ignore.case=TRUE)]
b20 <- u[grepl("Volcanic Ash",u,ignore.case=TRUE)]
b21 <- u[grepl("Waterspout",u,ignore.case=TRUE)]
b22 <- u[grepl("WildFire",u,ignore.case=TRUE)]
b23 <- u[grepl("Winter Storm",u,ignore.case=TRUE)]
b24 <- u[grepl("Winter Weather",u,ignore.case=TRUE)]
# 

x_filtered <- x[x$EVTYPE %in% c(a1,a2,a3,a4,a5,a6,a7,a8,a9,a10,a11,a12,a13,a14,a15,a16,a17,a18,
                                a19,a20,a21,a22,a23,a24,b1,b2,b3,b4,b5,b6,b7,b8,b9,b10,b11,b12,b13,
                                b14,b15,b16,b17,b18,b19,b20,b21,b22,b23,b24),]

v <- x_filtered$EVTYPE
v[x_filtered$EVTYPE %in% a1] = "Astromomical Low Tide"
v[x_filtered$EVTYPE %in% a2] = "Avalanche"
v[x_filtered$EVTYPE %in% a3] = "Blizzard"
v[x_filtered$EVTYPE %in% a4] = "Coastal Flood"
v[x_filtered$EVTYPE %in% a5] = "Cold/Wind Chill"
v[x_filtered$EVTYPE %in% a6] = "Debris Flow"
v[x_filtered$EVTYPE %in% a7] = "Dense Fog"
v[x_filtered$EVTYPE %in% a8] = "Dense Smoke"
v[x_filtered$EVTYPE %in% a9] = "Drought"
v[x_filtered$EVTYPE %in% a10] = "Dust Devil"
v[x_filtered$EVTYPE %in% a11] = "Dust Storm"
v[x_filtered$EVTYPE %in% a12] = "Excessive Heat"
v[x_filtered$EVTYPE %in% a13] = "Extreme Cold/Wind Chill"
v[x_filtered$EVTYPE %in% a14] = "Flash Flood"
v[x_filtered$EVTYPE %in% a15] = "Flood"
v[x_filtered$EVTYPE %in% a16] = "Frost/Freeze"
v[x_filtered$EVTYPE %in% a17] = "Funnel Cloud"
v[x_filtered$EVTYPE %in% a18] = "Freezing Fog"
v[x_filtered$EVTYPE %in% a19] = "Hail"
v[x_filtered$EVTYPE %in% a20] = "Heat"
v[x_filtered$EVTYPE %in% a21] = "Heavy Rain"
v[x_filtered$EVTYPE %in% a22] = "Heavy Snow"
v[x_filtered$EVTYPE %in% a23] = "High Surf"
v[x_filtered$EVTYPE %in% a24] = "High Wind"
v[x_filtered$EVTYPE %in% b1] = "Hurricane/Typhoon"
v[x_filtered$EVTYPE %in% b2] = "Ice Storm"
v[x_filtered$EVTYPE %in% b3] = "Lake-Effect Snow"
v[x_filtered$EVTYPE %in% b4] = "Lakeshore Flood"
v[x_filtered$EVTYPE %in% b5] = "Lightning"
v[x_filtered$EVTYPE %in% b6] = "Marine Hail"
v[x_filtered$EVTYPE %in% b7] = "Marine High Wind"
v[x_filtered$EVTYPE %in% b8] = "Marine Strong Wind"
v[x_filtered$EVTYPE %in% b9] = "Marine Thunderstorm Wind"
v[x_filtered$EVTYPE %in% b10] = "Rip Current"
v[x_filtered$EVTYPE %in% b11] = "Seiche"
v[x_filtered$EVTYPE %in% b12] = "Sleet"
v[x_filtered$EVTYPE %in% b13] = "Storm Surge/Tide"
v[x_filtered$EVTYPE %in% b14] = "Strong Wind"
v[x_filtered$EVTYPE %in% b15] = "Thunderstorm Wind"
v[x_filtered$EVTYPE %in% b16] = "Tornado"
v[x_filtered$EVTYPE %in% b17] = "Tropical Depression"
v[x_filtered$EVTYPE %in% b18] = "Tropical Storm"
v[x_filtered$EVTYPE %in% b19] = "Tsunami"
v[x_filtered$EVTYPE %in% b20] = "Volcanic Ash"
v[x_filtered$EVTYPE %in% b21] = "Waterspout"
v[x_filtered$EVTYPE %in% b22] = "Wildfire"
v[x_filtered$EVTYPE %in% b23] = "Winter Storm"
v[x_filtered$EVTYPE %in% b24] = "Winter Weather"

```
### The final filtered database (x_filtered) only contain 39 of the 48 weather event categories defined above. The  total number of fatalities and injuries across the data sampling period was summed to give the number of casualties for each weather category.

```{r, Weather-related casualty data,echo=TRUE}
x_filtered$Event_Categories <- v
u <- unique(x_filtered$Event_Categories)
y <- matrix(0,length(u),2)
for (j in 1:length(u)) {
  y[j,] <- colSums(x_filtered[x_filtered$Event_Categories == u[j],c("FATALITIES","INJURIES")],na.rm=TRUE)
}
  
index <- order(as.numeric(rowSums(y)),as.numeric(y[,2]),as.numeric(y[,1]),decreasing=TRUE)
df <- data.frame(EVENT=u[index],FATALITIES=as.numeric(y[index,1]),
    INJURIES=as.numeric(y[index,2]),CASUALTY_COUNT=as.numeric(rowSums(y[index,])),
    PERCENT_OF_TOTAL=100*as.numeric(rowSums(y[index,]))/sum(as.numeric(rowSums(y))))
```

### The casualty data arising from all the monitored weather events by event category, was estimated  and the results displayed in the the table below (TABLE 1). Note that only 31 of the 39 weather categories contained casualty data.

```{r, echo=FALSE}
head(df,31)
```

### An event severity index for each weather event was determined by taking the log of all casualty numbers for, i.e. 

```{r, Event severity index estimation, data,echo=TRUE}
event_severity_index <- log(df$CASUALTY_COUNT[1:31])
```

###The following plot tracks the event severity index by event category, where the even categories are displayed from the most consequential to the least consequential:

```{r,Event severity index versus weather event,echo=TRUE}
par(mar = c(7, 4, 4, 2) + 0.1) ## Increase bottom margin to make room for rotated labels
plot(event_severity_index[1:31], xaxt = "n", pch=19,cex=1.2,
     type=  "b",
     xlab = "",
     ylab="Event Severity Index",
      main="Severity of Human Casualties by Weather Event")
axis(1, labels = FALSE)
labels <- as.character(df$EVENT[1:31])
text(1:31,par("usr")[3] - 0.25, srt = 45, adj = 1,
     labels = labels, xpd = TRUE)
```

###Next, property and crop damage data were used to evaluate the ecomomic consequences of weather events. Specifically, data from each of the 50 states was evaluated and the weather event resulting in the highest monetary loss from property and crop damage was determined. The analysis was carried out as follows:

```{r,echo=TRUE}
xt = x_filtered # assign NOAA dataset to new variable xt and use the variables PROPDMGEXP and CROPDMGEXP to filter data
f1 <- union(union(grep("K",xt$PROPDMGEXP),grep("B",xt$PROPDMGEXP)),grep("M",xt$PROPDMGEXP))
f2 <- union(union(grep("K",xt$CROPDMGEXP),grep("B",xt$CROPDMGEXP)),grep("M",xt$CROPDMGEXP))
f <- union(f1,f2)
xt <- x_filtered[f,]
```

```{r, Initialize dataframe,echo=TRUE}
v <- unique(xt$STATE)
v <- c(v[1:3],"AK",v[4:49]) # 50 states
z <- data.frame(matrix(ncol = 3, nrow = length(v))) 
colnames(z) <- c("STATE","EVENT_CATEGORY","TOTAL_DAMAGES (TRILLION $)")
```

###  Property and crop damage amounts were retrieved for the weather event with the larget economic consequence for each state

```{r, Calculate and store property damage data,echo=TRUE}
for (j in 1:length(v)) {
  
  # Retrieve crop damage data
  select <- xt$STATE == v[j] & (xt$PROPDMG > 0 | xt$CROPDMG > 0)
  df <- xt[select,c("Event_Categories", "PROPDMG", "PROPDMGEXP","CROPDMG", "CROPDMGEXP")]
  df$PROPDMG <- as.numeric(df$PROPDMG)
  df$CROPDMG <- as.numeric(df$CROPDMG)
  df$PROPDMG[df$PROPDMGEXP == "K"] <- array(df$PROPDMG * 1000,length(df$PROPDMG[df$PROPDMGEXP == "K"]))
  df$PROPDMG[df$PROPDMGEXP == "M"] <- array(df$PROPDMG * 1000000,length(df$PROPDMG[df$PROPDMGEXP == "M"]))
  df$PROPDMG[df$PROPDMGEXP == "B"] <- array(df$PROPDMG * 1000000000,length(df$PROPDMG[df$PROPDMGEXP == "B"]))
  
  df$CROPDMG[df$CROPDMGEXP == "K"] <- array(df$CROPDMG * 1000,length(df$CROPDMG[df$CROPDMGEXP == "K"]))
  df$CROPDMG[df$CROPDMGEXP == "M"] <- array(df$CROPDMG * 1000000,length(df$CROPDMG[df$CROPDMGEXP == "M"]))
  df$CROPDMG[df$CROPDMGEXP == "B"] <- array(df$CROPDMG * 1000000000,length(df$CROPDMG[df$CROPDMGEXP == "B"]))
  
  # Calculate total property and crop damage for each event type in state j
  y2 <- matrix(0,length(u),1) 
  for (k in 1:length(u)) {
    y2[k] <- sum(colSums(df[df$Event_Categories == u[k],c("PROPDMG","CROPDMG")],na.rm=TRUE))
  }
  
  # Sort results to determine event with largest property damamge in state j
  select <- order(y2,u,decreasing=TRUE)

  # Store results in data frame with property damage amounts in Trillions
  z[j,] <- c(v[j],u[select[1]],y2[select[1]]/1000000000000)

}
```

### The data that ties the largest economic losses to a specific weather category across all 50 states (US Territories exluded), is summarized in the following table (TABLE2):

```{r,echo=FALSE}
z
```

###The table above indicates that only 6 of the 48 weather categories listed by NOAA (i.e. Flood, Thunderstorm Wind, Tornado, Hurricane/Typhoon, Hail and Storm Surge/Tide), are responsible for most of the economic losses in all 50 states. The plot below describes the number of states with the highest economic losses from each of the 6 weather categories:

```{r,echo=TRUE}
# Distribution of event-categories leading to the most property damages across all 50 states
event_count <- c(length(z[z$EVENT_CATEGORY=="Flood",2]),
                 length(z[z$EVENT_CATEGORY=="Thunderstorm Wind",2]),
                 length(z[z$EVENT_CATEGORY=="Tornado",2]),
                 length(z[z$EVENT_CATEGORY=="Hurricane/Typhoon",2]),
                 length(z[z$EVENT_CATEGORY=="Hail",2]),
                 length(z[z$EVENT_CATEGORY=="Storm Surge/Tide",2]))

vr <- c("Flood","Thunderstorm Wind","Tornado",
  "Hurricane/Typhoon","Hail","Storm Surge/Tide")


par(mar = c(7, 4, 4, 2) + 0.1) ## Increase bottom margin to make room for rotated labels
plot(event_count, xaxt = "n", pch=19,cex=1.5,
     type=  "b",
     xlab = "",
     ylab="Number of States",
    main = "Count of states with highest economic\nlosses from each weather category")
axis(1, labels = FALSE)
text(1:length(event_count),par("usr")[3] - 1, srt = 45, adj = 1,
     labels = vr, xpd = TRUE)

```

##Results:
###The analysis shows that 76% of all weather-related casualties from 1950 - 2011 are caused by weather activity that are linked to Tornados and Heat. Table 1 indicates that Tornados are the largest contributor to the casualty levels from weather related activity (64%). Tornados coupled with Heat and Thunderstorm wind are the top three weather events with respect to casualty levels - these account for 79% of all casualties. As for the ecomomic consequences of bad weather, data from the plot above shows flood accounts for the greatest amount of economic loss in 25 of the 50 states (50%). While Tornados are responsible for most casualties, they also the largest source of economic loss from weather related activity in 16 of 50 states (32%). Looking at both casualty data and economic loss data, Tornados are by far the single most distructive type of weather event.


