An Evaluation of Severe Weather Data from 1950 - 2011
==============================================================================
## Synopsis: 
### This report evaluates weather data acquired over 60 years (1950-2011) by the National Oceanic and Atmospheric Administration (NOAA). The analysis is narrowly focused on determining (1) the weather events that are most harmful to human health and (2) which weather events have the greatest ecomonic consequences across the United States. The analysis shows that 20 of the 985 unique weather events recorded over the measurement period account for 95% of all weather-related casualties (including injuries and fatalities) in the United States. Furthermore, 11 specific weather events are the most consequential with respect to property and crop damage across all 50 states.

## Data Processing: 
###The NOAA database was downloaded from web via the John Hopkins Coursera portal and stored on the computer with which the analysis was done. The .csv file (repdata-data-StormData.csv) was then loaded in the R environment as follows:

```{r, echo=TRUE}
setwd("C://Users//solasoy//Dropbox//R//Coursera//Reproducible Research")
x <- read.csv(file='repdata-data-StormData.csv',
              colClasses="character")
```
###Next, for each of the 985 unique weather events, the number of total number of fatalities and injuries across the data sampling period was summed to give the number of casualties stemming from this weather activity.

```{r, Weather-related casualty data,echo=TRUE}
x$INJURIES <- as.numeric(x$INJURIES)
x$FATALITIES <- as.numeric(x$FATALITIES)

u <- unique(x$EVTYPE) # unique weather events
y <- matrix(0,length(u),2)
for (j in 1:length(u)) {
  y[j,] <- colSums(x[x$EVTYPE == u[j],c("FATALITIES","INJURIES")],na.rm=TRUE)
}
  
index <- order(as.numeric(rowSums(y)),as.numeric(y[,2]),as.numeric(y[,1]),decreasing=TRUE)
df <- data.frame(EVENT=u[index],FATALITIES=as.numeric(y[index,1]),
    INJURIES=as.numeric(y[index,2]),CASUALTY_COUNT=as.numeric(rowSums(y[index,])),
    PERCENT_OF_TOTAL=100*as.numeric(rowSums(y[index,]))/sum(as.numeric(rowSums(y))))
```

###An event severity index for each weather event was determined by taking the log of all casualty numbers for, i.e. 

```{r, Event severity index estimation, data,echo=TRUE}
event_severity_index <- log(df$CASUALTY_COUNT)
```

### The casualty data arising from all the monitored weather events were captured in a data frame. The top 20 weather events with respect to casualty levels, are as follows:

```{r, echo=TRUE}
head(df,20)
```

###The following plot tracks the event severity index by weather type, and shows the most consequential weather events:

```{r,Event severity index versus weather event,echo=TRUE}
par(mar = c(7, 4, 4, 2) + 0.1) ## Increase bottom margin to make room for rotated labels
plot(event_severity_index[1:20], xaxt = "n", pch=19,
     type=  "b",
     xlab = "",
     ylab="Event Severity Index",
     main="Top 20 most harmful weather events to humans")
axis(1, labels = FALSE)
labels <- df$EVENT[1:20]
text(1:20,par("usr")[3] - 0.25, srt = 45, adj = 1,
     labels = labels, xpd = TRUE)
```

###The following plot highlights the percentage of all weather-related casualties in the last 60 years that derive from each of the top 20 weather events:

```{r, Percent of total casualties versus weather event,echo=TRUE}
par(mar = c(7, 4, 4, 2) + 0.1) 
with(df,plot(PERCENT_OF_TOTAL[1:20],pch=19,
      ylab="Percent of Total Casualties",
      xlab=NA,
      type="b", xaxt = "n",
      main="Weather events responsible for 94.6%\nof total number of casualties"))

axis(1, labels = FALSE)
labels <- df$EVENT[1:20]
text(1:20,par("usr")[3] - 0.25, srt = 45, adj = 1,
     labels = labels, xpd = TRUE)
```

###Next, property and crop damage data were used to evaluate the ecomomic consequences of weather events. Specifically, data from each of the 50 states was evaluated and the weather event resulting in the highest monetary loss from property and crop damage was determined. The analysis was carried out as follows:

```{r, Pre-filter Data,echo=TRUE}
xt = x # assign NOAA dataset to new variable xt and use the variables PROPDMGEXP and CROPDMGEXP to filter data
f1 <- union(union(grep("K",xt$PROPDMGEXP),grep("B",xt$PROPDMGEXP)),grep("M",xt$PROPDMGEXP))
f2 <- union(union(grep("K",xt$CROPDMGEXP),grep("B",xt$CROPDMGEXP)),grep("M",xt$CROPDMGEXP))
f <- union(f1,f2)
xt <- x[f,]
```

```{r, Initialize dataframe,echo=TRUE}
v <- unique(xt$STATE)
v <- c(v[1:3],"AK",v[4:49]) # 50 states
z <- data.frame(matrix(ncol = 3, nrow = length(v))) 
colnames(z) <- c("STATE","EVENT","TOTAL_DAMAGES (TRILLION $)")
```

### Retrieve property and crop damage amounts for the event with the larget economic consequence for each state

```{r, Calculate and store property damage data,echo=TRUE}
for (j in 1:length(v)) {
  
  # Retrieve crop damage data
  select <- xt$STATE == v[j] & (xt$PROPDMG > 0 | xt$CROPDMG > 0)
  df <- xt[select,c("EVTYPE", "PROPDMG", "PROPDMGEXP","CROPDMG", "CROPDMGEXP")]
  df$PROPDMG <- as.numeric(df$PROPDMG)
  df$CROPDMG <- as.numeric(df$CROPDMG)
  df$PROPDMG[df$PROPDMGEXP == "K"] <- array(df$PROPDMG * 1000,length(df$PROPDMG[df$PROPDMGEXP == "K"]))
  df$PROPDMG[df$PROPDMGEXP == "M"] <- array(df$PROPDMG * 1000000,length(df$PROPDMG[df$PROPDMGEXP == "M"]))
  df$PROPDMG[df$PROPDMGEXP == "B"] <- array(df$PROPDMG * 1000000000,length(df$PROPDMG[df$PROPDMGEXP == "B"]))
  
  df$CROPDMG[df$CROPDMGEXP == "K"] <- array(df$CROPDMG * 1000,length(df$CROPDMG[df$CROPDMGEXP == "K"]))
  df$CROPDMG[df$CROPDMGEXP == "M"] <- array(df$CROPDMG * 1000000,length(df$CROPDMG[df$CROPDMGEXP == "M"]))
  df$CROPDMG[df$CROPDMGEXP == "B"] <- array(df$CROPDMG * 1000000000,length(df$CROPDMG[df$CROPDMGEXP == "B"]))
  
  # Calculate total property and crop damage for each event type in state j
  y2 <- matrix(0,length(u),1) 
  for (k in 1:length(u)) {
    y2[k] <- sum(colSums(df[df$EVTYPE == u[k],c("PROPDMG","CROPDMG")],na.rm=TRUE))
  }
  
  # Sort results to determine event with largest property damamge in state j
  select <- order(y2,u,decreasing=TRUE)

  # Store results in data frame with property damage amounts in Trillions
  z[j,] <- c(v[j],u[select[1]],y2[select[1]]/1000000000000)

}
```

### The data that ties the largest economic losses to the a specific type of weather event in all of the 50 states, is summarized in the following table:

```{r,echo=FALSE}
z
```

##Results:
###The analysis shows that 94.6% of all weather-related casualties from 1950 - 2011 are caused by 20 of the 985 types of weather events that are tracked by NOAA. As for the ecomomic consequences of bad weather, the following 11 weather events: flash floods, thunderstorm wind, tornados, floods, winter storms, hurricane/typhoons, high winds, river floods, storm surges, hail and hurricanes account for the most amount of property and/or crop damage in all 50 states.