Assignment 1
=================================================================================================

Data from a personal monitoring device was downloaded as follows:

```{r,echo=TRUE}
setwd("C://Users//solasoy//Dropbox//R//Coursera//Reproducible Research")
x <- read.csv('activity.csv')
```

# Q1: What is mean total number of steps taken per day?

## Total number of steps taken per day

```{r,echo=TRUE,result="asis"}
grp <- split(as.numeric(x$steps),x$date)
spd <- sapply(grp,sum)
data.frame(steps_per_day=spd)
```

## Histogram of the total number of steps taken each day

```{r,echo=TRUE}
hist(spd[!is.na(spd)],
     breaks=12,
     col="blue",
     main = "Histogram of Steps_Per_Day",
     xlab = "Total Steps_Per_Day")
```

## Mean and median of the total number of steps taken per day

```{r,echo=TRUE,result="asis"}
mean(spd[!is.na(spd)])
median(spd[!is.na(spd)])
```

# Q2: What is the average daily activity pattern?

## Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)


```{r,echo=TRUE}
n <- split(as.numeric(x$steps[!is.na(x$steps)]),x$interval[!is.na(x$steps)])
avg_steps <- sapply(n,mean)
interval = unique(x$interval)

plot(interval,avg_steps,type="l",
     xlab="5min Interval",
     ylab= "Steps averaged across all days")
```

## 5-minute interval on average across all the days in the dataset, containing the maximum number of steps

```{r,echo=TRUE,result="asis"}
interval[avg_steps == max(avg_steps)]
```

# Q3: Imputing missing values

## Total number of missing values in the dataset 

```{r,echo=TRUE,result="asis"}
length(x[is.na(x[,1]),1])
```

## Determine distribution of missing values in dataset

```{r,echo=TRUE}
v <- unique(x$date)
h <- seq(1,length(v))
for (k in 1:length(h)) h[k] <- length(x[is.na(x$steps) & x$date==v[k],1])
plot(h,type="b",xlab="Day Number",ylab="Number of Missing Values")
```

### The plot above indicates that the missing values are limited to only 8 of the 61 days of this experiment. On each of the 8 days, no steps were recored i.e. all  288 measurements are missing


## Create new dataset with missing values replaced with median value of steps at each interval across all days

```{r,echo=TRUE}
median_steps <- sapply(n,median)
x_fill = x
for (j in 1:length(v)) {
  if (length(x_fill[is.na(x_fill$steps) & x_fill$date==v[j],1])==288) {
    x_fill[is.na(x_fill$steps) & x_fill$date==v[j],1] <- median_steps 
  } 
}
```

## Generate histogram of total number of steps from new dataset and compare with original

```{r,echo=TRUE}
grp_fill <- split(x_fill$steps,x_fill$date)
spd_fill <- sapply(grp_fill,sum)

hist(spd_fill,
     breaks=12,
     col="blue",
     main = "Histogram of Steps_Per_Day (filled data)",
     xlab = "Steps_Per_Day")
```

### From the histogram above, we observe a larger frequency of total number of steps per day that is approximately 2000 or less, relative to the initial histogram; otherwise the frequency distribution for this histogram is similar to the first one.

## Mean and median total number of steps taken per day from new dataset

```{r,echo=TRUE,result="asis"}
mean(spd_fill)
median(spd_fill)
```

# Q4: Are there differences in activity patterns between weekdays and weekends?

## Create a new factor variable in the dataset with two levels - "weekday" and "weekend"

```{r,echo=TRUE}
u <- weekdays(as.Date(x_fill$date))
h <- unique(x_fill$date)
tarray <- seq(as.POSIXct(h[1]), as.POSIXct(h[length(h)]),"days")
w <- weekdays(tarray)
day_flag <- array(0,length(u))
for (j in 1:length(w)) {
  if (w[j] =="Saturday" | w[j]=="Sunday") day_flag[u==w[j]] = "weekend"
}
day_flag [day_flag==0]="weekday"
x_new <- data.frame(cbind(x_fill,day_flag))
```

## Generate panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r,echo=TRUE}
library(lattice)

avg <- array(0,dim(x_new)[1])
for (i in 1:length(interval)) {
  avg[x_new$interval==interval[i] & x_new$day_flag == "weekday"] <- 
    mean(x_new[x_new$interval==interval[i] & x_new$day_flag == "weekday","steps"])
  
  avg[x_new$interval==interval[i] & x_new$day_flag == "weekend"] <- 
    mean(x_new[x_new$interval==interval[i] & x_new$day_flag == "weekend","steps"])               
}

df <- data.frame(interval=x_new$interval,
      average_steps=as.numeric(avg),
      day_flag=x_new$day_flag)

xyplot(average_steps ~ interval | day_flag,type = "l",df)
```